\pagenumbering{roman}

\begin{abstract} 

Variational autoencoders (VAEs) are widely used for generative modeling and representation learning tasks. This thesis presents two novel approaches aimed at enhancing the performance of VAEs through the integration of semi-conditional variational autoencoders (SCVAEs). The integration of SCVAEs and VAEs is motivated by the potential for improvement in the effectiveness of capturing the underlying data distribution and the improvement in generating high-quality samples.


The first method extends the traditional VAEs by incorporating a second conditioned decoder, thereby enabling the model to multi-task and learn better latent representations. The second method utilizes a unified decoder for both tasks by employing sophisticated training strategies. These approaches are implemented and evaluated on Gaussian VAEs and VQ-VAEs.

Extensive experimentation is conducted across diverse image datasets, including MNIST, CIFAR10, and CelebA. The results show that, in certain cases, the proposed methods yield superior performance compared to standard VAE architectures. By bridging the gap between SCVAEs and VAEs, this work gives new insights on how to improve the methods further and opens up new avenues for future research.

\end{abstract}

\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}

	First, I would like to express my deepest gratitude to my supervisors, Nello Blaser and Kristian Gundersen, for their unwavering support and helping to stay on track throughout the master's project. During this project, their mentorship has really helped to shape the work and improve the quality of this thesis. I am especially thankful for their generosity in sharing their time and insights during our meetings. The discussions we've had have provided invaluable clarity and direction to my work. 

	I am also grateful to my friends and family for their unwavering support, even during my time abroad. Their encouragement and understanding not only helped me to navigate the challenges of academic life but also played a crucial role in preventing homesickness. 

	Special thanks to my fiancée for her endless patience and unwavering support throughout this journey.
	
	\vspace{1cm}
	\hspace*{\fill}\texttt{Edvards Zakovskis}\\ 
	\hspace*{\fill}\today
\end{abstract}
\setcounter{page}{1}
\newpage